
# Statistical file formats

## SPSS, SAS and Stata

### haven

> Haven enables R to read and write various data formats used by other statistical packages by wrapping the fantastic [ReadStat](http://www.evanmiller.org/) C library written by Evan Miller. Haven is part of the [tidyverse](http://tidyverse.org/).
> 
> --- https://haven.tidyverse.org/

#### Pros {-}

* Labelled datasets!

* Good support for recent file formats, and good translation of data types into appropriate R classes.

* Supports writing as well as reading.

#### Cons {-}

* For the non-tidyverse fans - deeply embedded in the tidyverse way of doing things.

* Somewhat stable, but still has the occasional breaking change.

#### Examples {-}

The example below reads the GSS dataset in both SPSS and Stata formats.

Note the inclusion of the `user_na = TRUE` for reading SPSS files. By default `read_sav()` converts user tagged `NA` values to `NA` in R - setting `user_na = TRUE` retains these values. We'll go into more detail on working with tagged missing values in section \@ref(missing-values).

```{r, eval = FALSE}
install.packages("haven")
library(haven)

gss <- read_sav("data/gss/GSS2018.sav", user_na = TRUE)

gss <- read_stata("data/gss/GSS2018.dta")
```

### foreign

> Reading and writing data stored by some versions of 'Epi Info', 'Minitab', 'S', 'SAS', 'SPSS', 'Stata', 'Systat', 'Weka', and for reading and writing some 'dBase' files.
> 
> --- https://cran.r-project.org/package=foreign

#### Pros {-}

* Extremely stable.

* Supported and developed by the R Core Team.

* Supports additional file formats not supported by haven.

#### Cons {-}

* Splits long character variables into 255 character variables.

* Inconsistent support for newer file formats (e.g. no support for Stata after version 12).

* Difficult to use categorical labels without converting to factors.

#### Examples {-}

The example below reads the GSS dataset in both SPSS and Stata formats.

```{r, eval = FALSE}
install.packages("foreign")
library(foreign)

gss <- read.spss("data/gss/GSS2018.sav", use.value.labels = FALSE) %>%
  as_tibble()

gss <- read.dta("data/gss/GSS2018.dta", convert.factors = FALSE) %>%
  as_tibble()
```

## Excel

No discussion of survey datasets would be complete without a mention of good old Excel.
Our advice is to avoid Excel if possible - Excel files can have strange import issues, particularly with date formats and formulas.

Where it is necessary to import Excel files there are many choices, but we would recommend these packages:

* [openxlsx](https://ycphs.github.io/openxlsx/) - a custom Excel library. In our experience this package has the least weird Excel issues, and is fantastic for writing and styling output tables.

* [readxl](https://readxl.tidyverse.org/) - a custom Excel library, part of the tidyverse. Very fast and reliable, but can do unexpected things when trying to guess column types. Does not support writing Excel workbooks.

* [xlsx](https://github.com/colearendt/xlsx) - a wrapper for the [Apache POI](https://poi.apache.org/) Java library. Apache POI is well maintained, but extremely high memory usage for larger datasets.
