--- 
title: "Survey Research Datasets and R"
author: "Danny Smith"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: ["book.bib", "packages.bib"]
biblio-style: apalike
link-citations: yes
url: 'https\://socialresearchcentre.github.io/r_survey_datasets/'
github-repo: 'socialresearchcentre/r_survey_datasets'
---

```{r setup, include=FALSE}
options(
  htmltools.dir.version = FALSE, formatR.indent = 2,
  width = 80, digits = 4, warnPartialMatchAttr = FALSE, warnPartialMatchDollar = FALSE
)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

## Abstract

Although R began as a specialist statistical programming language, the R ecosystem has grown wildly over the past few years making it a viable general-purpose research environment across the whole research lifecycle.

Survey research datasets come from a diverse range of sources, often containing richer metadata than your average data frame. This workshop provides a practical demonstration of several packages for accessing and working with survey data, associated metadata and official statistics in R.

We will demonstrate:

* Working with external data sources from common statistical packages (SPSS, SAS, Stata, Excel) and their quirks

* Easily working with categorical data in R with the “labelled” R package

* A brief introduction to the “tidy data” concept (https://tidyr.tidyverse.org/articles/tidy-data.html) and its application to survey research data

* Accessing external databases in an R native way using DBI and dbplyr

* Accessing publicly accessible data in R scripts via the web

* Resources for accessing official statistics data in R

Participants should have a basic working knowledge of R to follow along with examples, but beginners are also welcome.

## Datasets

[American Trends Panel Wave 56](https://www.pewsocialtrends.org/dataset/american-trends-panel-wave-56/)

[Teens and Tech Survey 2018](https://www.pewresearch.org/internet/dataset/teens-and-tech-survey-2018/)

# Statistical file formats

## Excel

[openxlsx](https://ycphs.github.io/openxlsx/)

[xlsx](https://github.com/colearendt/xlsx)

[readxl](https://readxl.tidyverse.org/)

## foreign

https://cran.r-project.org/web/packages/foreign/index.html

## haven

[haven](https://haven.tidyverse.org/)

```{r}
library(haven)


```

# Labelled data

https://larmarange.github.io/labelled/

## Definitions

### Variable labels

### Value labels

### Missing values



## Usage

### Exploring datasets

### Interactive printing



### Tabling

* gtsummary

https://www.danieldsjoberg.com/gtsummary/

# Accessing databases

```{r}
library(dplyr, warn.conflicts = FALSE)

con <- DBI::dbConnect(RSQLite::SQLite(), ":memory:")
copy_to(con, mtcars)
```

## DBI

https://dbi.r-dbi.org/

## dbplyr

https://dbplyr.tidyverse.org/

# Official statistics

https://github.com/SNStatComp/awesome-official-statistics-software
